\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2022-01-31 manuscript/paper.tex
Update the likelihood

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that we include $\lambda_0$ in $\theta$. Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time. So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to the number of unique failure times (denoted $J$ thereafter).
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that the random effects vector $b_i$, the group membership account for the association between the longitudinal and event outcomes and the association features $\Psi_i$, that is
\begin{equation}
  \label{eq:ind-hyp}
  f(t_i, \delta_i, y_i | b_i, G_i = k, \Psi_i ; \theta) = f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(y_i | G_i = k)\int_{\R^r} f(t_i, \delta_i| b_i, G_i = k ; \theta) f(b_i | y_i, G_i = k; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\}. \]

@@

In order to derive an algorithm for this objective, we introduce a so-called prox-QNMCEM algorithm, being a combination between an EM algorithm~\citep{dempster1977maximum} with Monte Carlo approximations~\citep{levine2001implementations}, an L-BFGS-B algorithm~\citep{zhu1997algorithm} and a proximal gradient descent algorithm~\citep{beck2009fast}. EM algorithm has already been used for multivariate data joint modeling (see~\citet{lin2002maximum} for instance), but here we face different original problems: for each subject $i$, the latent variables are the pairs $(G_i, b_i)$ (not only the random effects); and then, we want to minimize the penalized objective $\ell_n^\text{pen}$ (not ``only'' the negative log-likelihood).

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_{k,i}(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_{k,i}(s) \big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

>>>>

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that we include $\lambda_0$ in $\theta$. Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time. So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to the number of unique failure times (denoted $J$ thereafter).
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that the random effects vector $b_i$, the group membership account for the association between the longitudinal and event outcomes and the association features $\Psi_i$, that is
\begin{equation}
  \label{eq:ind-hyp}
  f(t_i, \delta_i, y_i | b_i, G_i = k, \Psi_i ; \theta) = f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) \int_{\R^r} f(t_i, \delta_i| b_i, G_i = k, \Psi_i ; \theta) f(y_i | b_i, G_i = k, \Psi_i; \theta) f(b_i ; \theta) \dd b_i \\
  &= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | G_i = k; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) = \big[\lambda(t_i|G_i = k, \Psi_i)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|G_i = k, \Psi_i) \dd s \Big\}. \]

@@

In order to derive an algorithm for this objective, we introduce a so-called prox-QNEM algorithm, being a combination between an EM algorithm~\citep{dempster1977maximum}, an L-BFGS-B algorithm~\citep{zhu1997algorithm} and a proximal gradient descent algorithm~\citep{beck2009fast}. EM algorithm has already been used for multivariate data joint modeling (see~\citet{lin2002maximum} for instance), but here we face different original problems: for each subject $i$, the latent variables are the pairs $(G_i, b_i)$ (not only the random effects); and then, we want to minimize the penalized objective $\ell_n^\text{pen}$ (not ``only'' the negative log-likelihood).

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i \Big) - \exp \big\{ \gamma_k^\top \Psi_i \big\} \int_0^{t_i} \lambda_0(s) \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.
